---
title: "R for Stata Users"
author: "Purushottam Mohanty"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  xaringan::moon_reader:
    css:
    - default
    - metropolis
    - metropolis-fonts
    nature:
      highlightStyle: github
      highlightLines: yes
      countIncrementalSlides: no
  pdf_document: default
subtitle: '02: Data Wrangling'
---

```{css, echo=FALSE}
# CSS for including pauses in printed PDF output (see bottom of lecture)
@media print {
  .has-continuation {
    display: block !important;
  }
}

# stata code chunk background colour
.code-bg-stata .remark-code, .code-bg-stata .remark-code * {
 background-color:Gainsboro!important;
}

# count for a particular slide
.hide-count .remark-slide-number {
  display: none;
}

```


```{r setup, include=FALSE}

options(htmltools.dir.version = FALSE)

if (!require(pacman)) install.packages('pacman', repos = 'https://cran.rstudio.com')
pacman::p_load(knitr, tidyverse, pagedown, kableExtra, formatR)

opts_chunk$set(
  fig.align="center",  
  fig.height=4, #fig.width=6,
  # out.width="748px", #out.length="520.75px",
  dpi=300, #fig.path='Figs/',
  cache=T#, echo=F, warning=F, message=F
  )

# save the built-in output hook
hook_output <- knitr::knit_hooks$get("output")

# set a new output hook to truncate text output
knitr::knit_hooks$set(output = function(x, options) {
  if (!is.null(n <- options$out.lines)) {
    x <- xfun::split_lines(x)
    if (length(x) > n) {
      # truncate the output
      x <- c(head(x, n), "....\n")
    }
    x <- paste(x, collapse = "\n")
  }
  hook_output(x, options)
})

```

# overview

Data wrangling or cleaning is an important aspect of any project and it is important to be well acquainted with all the tools provided by R to facilitate an easy transition from Stata to R.   
    
While Stata has specific commands for specific data cleaning operations, R is more versatile and often the same operation can be performed in multiple ways through different packages.           
      
In this slide deck I use `dplyr` and `tidyr` packages to perform data manipulation operations. The same tasks can also be performed using `baseR` functions but they are often more complicated and cumbersome.    
    
I also provide equivalent commands in Stata where possible to help easily understand R functions.

---

# tidyverse

The `tidyverse` package in R is a composite of many different data manipulation, functional programming and data visualization packages. Check documentation about [tidyverse](https://www.tidyverse.org/packages/), [dplyr](https://dplyr.tidyverse.org) and [tidyr](https://tidyr.tidyverse.org) for more information.

```{r}
# only loads the primary tidyverse packages 
library(tidyverse) 
tidyverse_packages()
```

Packages like `ggplot2` for data visualization, `forcats` for handling categorical data and `stringr` for text manipulation enable so many possibilities. Packages within tidyverse can be separately loaded as well.

---

# pipe %>% operator

Many different packages in R, including `tidyr` and `dplyr` follow the pipe `%>%` operator syntax which makes way for clean looking code and saves considerable time by not having to specify the dataframe name everytime.

```{r, echo=F}
df = gapminder::gapminder
```

```{r, eval=F}
# both lines are equivalent
df %>% filter(!continent == 'Europe') %>% group_by(continent, year) %>% 
  summarize(mean_gdppc = mean(gdpPercap))
summarise(group_by(filter(df, !continent == 'Europe'), continent, year), mean_gdppc = mean(gdpPercap))
```

The first line can be read as, specifying the dataframe, filtering the rows and then grouping based on column names and the summarizing the gdpPercap variable. As you can see the first line is easier to read and logical in nature. With complicated and lengthy code, `%>%` operator becomes extremely handy.

---

class: inverse, center, middle
name: dplyr

# dplyr | tidyr

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---


# create new columns (1/n)

In R, new columns can be created using `dplyr::mutate()` function.
```{r, eval=F}
df %>% 
  mutate(pop_mn = pop / 1000000)
```

Note that `mutate()` creates a new column if dataframe `df` doesn't have a column with the specified namespace (i.e. `pop_mn`) or overwrites the existing column. So, `mutate()` is a substitute for Stata commands `generate` and `replace` depending on the namespace provided.  

.code-bg-stata[
```{r, eval=F}
# Stata equivalent
generate pop_mn = pop / 1000000
replace pop_mn = pop / 1000000
```
]

---
# create new column (2/n)

The original dataframe imported from the `gapminder` package.
```{r, eval=T}
df 
```


---
# create new column (3/n)

New column `pop_mn` has been created.
```{r, eval=T}
df %>% 
  mutate(pop_mn = pop / 1000000)
```

---
# create new column (4/n)

Let's overwite `pop_mn` where `pop_mn` is log of population. Note how the old `pop_mn` column is overwritten with log population values. Also, see how the `%>%` function allows us to perform multiple operations on the same dataframe.
```{r, eval=T}
df %>% 
  mutate(pop_mn = pop / 1000000) %>%
  mutate(pop_mn = log(pop))
```

---

# filter rows (1/n)

In R rows can be filtered using `dplyr::filter()` function.
```{r, eval=F}
df %>% 
  filter(continent == "Europe")
```
This keeps only those rows for which continent is Europe. Similarly, for keeping all rows outside continent Europe one can use the "!=" logical operation. Multiple conditions can also be specified using,
```{r, eval=F}
df %>% 
  filter(continent %in% c("Europe", "Africa"))
df %>% 
  filter(continent == "Europe" | continent == "Africa")
```
Both lines above are equivalent. Note how the `%in%` syntax makes code much more concise and readable.

.code-bg-stata[
```{r, eval=F}
# Stata equivalent
keep if continent == "Europe"
keep if continent == "Europe" | continent == "Africa"
```
]

---
# filter rows (2/n)

Only keeps the rows where continent is Europe.
```{r, eval=T}
df %>% 
  filter(continent == "Europe")
```

---
# filter rows (3/n)

Keeps all rows where continent is either Europe or Africa. (check no. of rows)
```{r, eval=T}
df %>% 
  filter(continent %in% c("Africa", "Europe"))
```


---

# filter columns (1/n)

In R, columns can be filtered using the `dplyr::select()` function. 
```{r, eval=F}
df %>% 
  select(continent)
df %>%
  select(country, continent)
```

Similarly, columns can be dropped using a `-` sign before the column name 
```{r, eval=F}
df %>% 
  select(-country, -continent)
```

Unlike Stata, for R both keeping and dropping columns is done using the same function. Also, one can use `select()` to order columns with or without dropping columns using the `select(columnA, columnB, everything())` syntax. The `everything()` function selects all columns not specified in `select()`.

.code-bg-stata[
```{r, eval=F}
# Stata equivalent
keep country continent 
drop country continent
```
]

---
# filter columns (2/n)
```{r, eval=T}
df %>% 
  dplyr::select(continent, country)
```

---
# filter columns (3/n)
```{r, eval=T}
df %>% 
  dplyr::select(-continent, -country)
```


---
# order rows (1/n)

In R, rows can be ordered is ascending or descending order using the `dplyr::arrange()` function.
```{r, eval=F}
df %>% 
  arrange(year, gdpPercap) # ascending order
df %>%
  arrange(desc(year), desc(gdpPercap)) # descending order
```

.code-bg-stata[
```{r, eval=F}
# Stata equivalent
sort year gdpPercap
gsort -year -gdpPercap # using gtools
```
]

---
# order rows (2/n)

```{r, eval=T}
df %>% 
  arrange(year, gdpPercap) # ascending order
```

---
# order rows (3/n)

```{r, eval=T}
df %>%
  arrange(desc(year), desc(gdpPercap)) # descending order
```

---
# distinct (1/n)

In R, duplicates can be removed from a dataframe using `dplyr::distinct()` function. The `.keep_all = T` option ensures that all columns are kept after removal of the duplicates.
```{r, eval=F}
df %>% 
  distinct(.keep_all = T)
```

Duplicates can also be removed from a particular column using,
```{r, eval=F}
df %>% 
  distinct(country, .keep_all = T)
```

The `distinct()` can also be used to view unique values for a column(s).
```{r, eval=F}
df %>% 
  distinct(country) # all countries present in df
```

.code-bg-stata[
```{r, eval=F}
# Stata equivalent
duplicates drop # for all rows
duplicates drop country, force # for a column
duplicates report country
```
]

---
# distinct (2/n)
```{r, eval=T}
df %>% 
  distinct(.keep_all = T)
```

---
# distinct (3/n)
```{r, eval=T}
df %>% 
  distinct(country, .keep_all = T)
```

---
# distinct (4/n)
```{r, eval=T}
df %>% 
  distinct(country) 
```

---
# summarize (1/n)

Summarize operates with `mutate()` at the backend and creates a new dataframe with specified columns based on the statistics specified. Note that `summarize()` and `summarise()` are equivalent to each other and can be interchanged. 
```{r, eval=F}
df %>% 
  summarize(mean_pop = mean(pop), median_gdppc = median(gdpPercap))
```

Summarise can be performed across multiple columns in combination with the `across()` function.
```{r, eval=F}
df %>% 
  summarize(across(c("pop","gdpPercap"), mean))
```

.code-bg-stata[
```{r, eval=F}
# Stata equivalent
collapse (mean) pop (median) gdpPercap
```
]

---
# summarize (2/n)

```{r, eval=T}
df %>% 
  summarize(mean_pop = mean(pop), median_gdppc = median(gdpPercap))
```

```{r, eval=T}
df %>% 
  summarize(across(c("pop","gdpPercap"), mean))
```

.code-bg-stata[
```{r, eval=F}
# Stata equivalent
bysort continent year: egen mean_pop = mean(pop)
bysort continent year: egen median_gdppc = median(gdpPercap)
collapse (mean) mean_pop median_gdppc, by(continent year)
```
]

---
# group operations (1/n)

Grouped row operations can be performed using the `dplyr::group_by()` function. 
```{r, eval=F}
df %>% 
  group_by(continent, year) %>% # grouping variables
  mutate(mean_pop = mean(pop)) # group wise operation to perform
```

The `dplyr` pipe operation implies that dataset is grouped as long as a separate `ungroup()` function is provided. It's a healthy practice to provide `ungroup()` function after the end of the grouped operation to avoid confusion.
```{r, eval=F}
df %>% 
  group_by(continent, year) %>% # grouping variables
  mutate(mean_pop = mean(pop)) %>% # group wise operation to perform
  ungroup() %>% # dataframe is now ungrouped
  mutate(mean_gdppc = mean(gdpPercap)) # ungrouped operation
```

.code-bg-stata[
```{r, eval=F}
# Stata equivalent
bysort continent year: egen mean_pop = mean(pop)
egen mean_gdppc = mean(gdpPercap)
```
]

---
# group operations (2/n)

```{r, eval=T, message=F, out.lines = 10}
df %>% 
  group_by(continent, year) %>% # grouping variables
  summarize(mean_pop = mean(pop), mean_gdppc = mean(gdpPercap)) # group wise operation
```

.code-bg-stata[
```{r, eval=F}
# Stata equivalent
bysort continent year: egen mean_pop = mean(pop)
bysort continent year: egen mean_gdppc = mean(gdpPercap)
collapse (mean) mean_pop mean_gdppc, by(continent year)
```
]

---
# reshape - long to wide (1/n)

Dataframes can be transformed from long to wide using the `tidyr::pivot_wider()` function. Here's how the long dataframe looks.

```{r, eval=T}
df 
```

---
# reshape - long to wide (2/n)

Transforming from long to wide.
```{r, eval=T, out.lines=10}
df_wide = df %>%
  pivot_wider(names_from = year, values_from = c("lifeExp", "pop", "gdpPercap"))
df_wide
```

.code-bg-stata[
```{r, eval=F}
# Stata equivalent
reshape wide lifeExp pop gdpPercap, i(continent country) j(year)
```
]

---
# reshape - wide to long (3/n)
Dataframes can be transformed from wide to long using the `tidyr::pivot_longer()` function. We can transform the dataframe created earlier by,


```{r, eval=T, out.lines=10}
df_wide %>%
  pivot_longer(cols = c(-continent, -country), names_to = c("type", "year"),
               names_sep = "_", values_to = "values")
```

.code-bg-stata[
```{r, eval=F}
# Stata equivalent
reshape long lifeExp pop gdpPercap, i(continent country) j(year)
```
]

---
# reshape - wide to long (4/n)

You might have noticed that the long data doesn't look like the original long data. The original data was partially long, so we'll have to convert it back to wide.
```{r, eval=T}
df_wide %>% # wide data
  pivot_longer(cols = c(-continent, -country), names_to = c("type", "year"),
               names_sep = "_", values_to = "values") %>% # long data
  pivot_wider(names_from = "type", values_from = "values") # original data
```


---
# merging dataframes (1/n)

Two dataframes can be joined using using keys with the `*_join()` suite of functions. Dataframe X and Y can be merged using, 
1. `inner_join(X, Y, by = "COL_NAME")` - includes all rows of X and Y.    
    
2. `left_join(X, Y, by = "COL_NAME")` - includes all rows of X        
    
3. `right_join(X, Y, by = "COL_NAME")` - includes all rows of Y       
    
4. `full_join(X, Y, by = "COL_NAME")` - includes all rows of X or Y       
      
.code-bg-stata[
```{r, eval=F}
# Stata equivalent
merge 1:1 COL_NAME using `Y` 
merge m:1 COL_NAME using `Y` 
merge 1:m COL_NAME using `Y` 
merge m:m COL_NAME using `Y`
keep if _merge == 3 # includes all rows of X and Y
```
]

---
# merging dataframes (2/n)

Let's use the approval ratings of US Presidents from `datasets::presidents`, which is pre-installed in R and merge it with the `gapminder::gapminder` data. The data is a quarterly time-series data so we perform some basic manipulation first.

```{r, eval=T, out.lines = 10}

df_presidents = data.frame(year = time(datasets::presidents),
                           rating = as.matrix(datasets::presidents)) %>%
  mutate(year = substr(year,1,4)) %>%
  group_by(year) %>%
  summarise(approval_rating = mean(rating, na.rm=T)) # rating by year

df_presidents

```

---
# merging dataframes (3/n)
```{r, eval=T, out.lines = 10}
df %>% mutate(year = as.character(year)) %>%
  left_join(df_presidents, by = "year")
```
Since the `datasets::presidents` only contained data from 1945 to 1974, merged values after 1974 are missing (coded as `NA`).

.code-bg-stata[
```{r, eval=F}
# Stata equivalent
tostring year, replace
merge 1:1 year using `df_presidents`
keep if _merge == 1 | _merge == 2 # same as left_join()
```
]

---
# appending rows

Dataframes in R can be appended using the `dplyr::bind_rows()` function. The dataframes are appended based on column names however unlike many other R funtions, it allows appending dataframes even when dataframes have different number of columns. In this regard, it works almost exactly like Stata's `append` command. 
```{r, eval=F}
df %>%
  filter(continent == "Europe") %>%
  bind_rows(df) # any dataframe can be provided
```

R objects are also often appended using the `baseR::rbind()` function. It also works similarly but the column length of the two dataframe must be identical. It can be used through the `rbind(DATAFRAME1, DATAFRAME2)` syntax.

.code-bg-stata[
```{r, eval=F}
# Stata equivalent
append using `Y`
```
]

---
# rename columns

There are a number of ways to rename columns in R. The easiest way is through the `dplyr::rename()` function which can be combined with the `%>%` operator. 
```{r, eval=F}
df %>%
  rename(country_name = country)
```

Similarly, multiple columns can be renamed using the following syntax,
```{r, eval=F}
df %>%
  rename(country_name = country, continent_name = continent)
```
 
Columns can also be renamed using the `baseR::names()` funtion. 
```{r, eval=F}
names(df)[1] = "country"
```

.code-bg-stata[
```{r, eval=F}
# Stata equivalent
rename (country continent) (country_name continent_name)
```
]

---

class: inverse, center, middle
name: data.table

# data.table

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---
# intro

Until now we've largely been using `dplyr` and `tidyr` (i.e. `tidyverse`) packages along with `baseR` where necessary. `data.table` is another data manipulation or wrangling package that displays the true power of R.    
    
Detailed documentation about data.table can be found at [CRAN](https://cran.r-project.org/web/packages/data.table/data.table.pdf) and further examples can be found at the data.table [vignette](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html).   
    
Unique advantages of `data.table` can be listed down as,
1. insane speed (see next slide)    
    
2. concise code         
    
3. no dependency (stable code)      
    
4. memory efficient


---
# speed insanity

```{r, message=FALSE, error=FALSE, out.lines = 4}
pacman::p_load(data.table, microbenchmark)

collapse_dplyr = function() {
  storms %>%
    group_by(name, year, month, day) %>% 
    summarize(wind = mean(wind), pressure = mean(pressure), category = first(category))
  }
storms_dt = as.data.table(storms)
collapse_dt = function() {
  storms_dt[, .(wind = mean(wind), pressure = mean(pressure), category = first(category)),
            by = .(name, year, month, day)]
  }
microbenchmark(collapse_dplyr(), collapse_dt(), times = 10)
```

Check the computation time difference between `data.table` and `dplyr`.

---
# data.table class

As mentioned earlier, every object in R has it's own class and functions can be applied to that object depending on the class of the object.    
    
`tidyverse` converts dataframes into a tibble (`tbl_df` class) in order to operate on them. Similarly, `data.table` functions can be applied to an object of `data.table` class. There are few ways to do that.   
    
1. `data.table(x = 1:20, y = 50:70)` - to create data.table from scratch        
    
2.  `as.data.table()` - convert existing dataframes or tibbles to data.table        
    
3. `setDT()` - reference existing dataframe as data.table. It does not create a copy of the data unlike `as.data.table()` hence need not be assigned to a new object.           
    
4. `fread("./example.csv")` - imports csv file extremely fast as a data.table object.   


---
# operations

The `data.table` function follows the following syntax `DT[i, j, by]` where **i** refers to row-wise operations, **j** refers to column wise-operations and **by** refers to group based operations.   

The `dplyr` equivalents of such operations are:
1. **i** - `filter()`, `arrange()`, `slice()`
2. **j** - `mutate()`, `select()`
3. **by** - `group_by()`, `add_count()`
    
For example,
```{r, eval=T, results='hide'}
dt_storms = as.data.table(storms)
dt_storms[status == "hurricane", mean(wind, na.rm=T), by = year]
```

.code-bg-stata[
```{r, eval=F}
# Stata equivalent
keep if status == "hurricane"
collapse (mean) wind, by(year)
```
]

---
# row operations (1/n)

```{r, eval=T, out.lines = 12}
dt_storms[status == "hurricane",]
```

Note that `dt_storms[status == "hurricane"]` (without comma) is equivalent to the above operation. 
.code-bg-stata[
```{r, eval=F}
# Stata equivalent
keep if status == "hurricane"
```
]

---
# row operations (2/n)

```{r, eval=T, out.lines = 12}
dt_storms[status == "hurricane" & year > 2000 & category > 1,]
```

Note that `dt_storms[status == "hurricane"]` (without comma) is equivalent to the above operation. 
.code-bg-stata[
```{r, eval=F}
# Stata equivalent
keep if status == "hurricane" & year > 2000 & category > 1
```
]

---
# column operations (1/n)

The `data.table` equivalent of `mutate()` is the `:=` operator which works with reference (i.e. it changes the original data.table so assignment is not required).    
      
New columns can be created using the,
- `dt_storms[, wind_scaled := wind / 100]`: creates new column `wind_scaled` from existing column `wind`.
- `dt_storms[, wind := wind^2]`: changes exisitng column `wind`

Note that results are not printed while make such changes. In order to print results, one has to mention `[]` after the end of the operation. For example,
```{r, eval=F}
dt_storms[, wind := wind^2][]
```

.code-bg-stata[
```{r, eval=F}
# Stata equivalent
generate wind_scaled = wind / 100
replace wind = wind^2
```
]

---
# column operations (2/n)

Columns can be removed from a data.table using the following syntax,
```{r, eval=F}
dt_storms[, ts_diameter := NULL]
```

Note that for memory efficiency data.table functions work by reference implying that changes made to a copy of a data are reflected in the original data as well. For example,
```{r, eval=F}
dt_storms = as.data.table(storms)
dt_new = dt_storms
dt_new[, ts_diameter := NULL] # also drop ts_diameter from dt_storms
```
This can be avoided by creating a true copy of the original data.table using `data.table::copy()`. For example,
```{r, eval=F}
dt_storms = as.data.table(storms)
dt_new = copy(dt_storms) 
```

.code-bg-stata[
```{r, eval=F}
# Stata equivalent
drop ts_diameter
```
]

---
# column operations (3/n)

In order to operate on **multiple columns** at once `data.table` provides two different options which perform identically.
```{r, eval=F}
DT[, c("newCol1","newCol2") := .(col1, col2)]
DT[, ':=' (newCol1=col1, newCol2=col2)]
```

Like `dplyr`, the `data.table` also allows the usage of the **pipe** `%>%` operator but a `.` needs to be added at the beginning of any operation.
```{r, eval=F}
DT %>%
  .[, newCol1 := col1] %>%
  .[, newCol2 := col2]
```

.code-bg-stata[
```{r, eval=F}
# Stata equivalent
generate newCol1 = col1
generate newCol2 = col2
```
]

---
# column operations (4/n)
New data.table can be created from scratch using the `data.table()` function. For example, 
```{r, eval=T}
dt = data.table(cities = c("London","Paris"), rent_eur = c(1800,1400))
dt
```

data.table works by reference so it changes original data.table without assignment.
```{r, eval=T}
dt[, rent_inr := rent_eur * 85] # references data.table (no assignment needed)
dt
```


---
# column operations (5/n)

Operations on a copy of data.table modifies the original data.table as well.
```{r, eval=T}
dt_new = dt
dt_new[, rent_eur := NULL]
dt # original data.table has changed
```

Operations on true copy of data.table doesn't modify original data.table.
```{r, eval=T}
dt = data.table(cities = c("London","Paris"), rent_eur = c(1800,1400))
dt_new = copy(dt)
dt_new[, cities := NULL]
dt # original data.table unmodified
```

---
# grouped operation (1/n)

Grouped operation in data.table is straightforward and similar to `dplyr::group_by()` and Stata's `bysort` command. For example to obtain summary statistics by group,
```{r, eval=F}
dt_storms[, mean(wind, na.rm = TRUE), by = .(status,category)]
```

In order to explicitly specify the new column name,
```{r, eval=F}
dt_storms[, .(mean_wind = mean(wind, na.rm = TRUE)), by = .(status,category)]
```

In order to add mean values to a new column rather without summarizing (collapsing) the data.table, we can use,
```{r, eval=F}
dt_storms[, mean_wind := mean(wind, na.rm = TRUE), by = .(status,category)]
```

.code-bg-stata[
```{r, eval=F}
# Stata equivalent
bysort status category: egen mean_wind = mean(wind)
collapse (mean) mean_wind, by(status category)
collapse (mean) wind, by(status category) # same as above 2 lines together
```
]

---
# grouped operation (2/n)
```{r, eval=T}
dt_storms[, .(mean_wind = mean(wind, na.rm = TRUE)), by = .(status,category)]
```

```{r, eval=T}
dt_storms[!status == "tropical storm", .(mean_wind = mean(wind, na.rm = TRUE)),
          by = .(status,category)]
```

---
# data.table features

data.table has a number of other functions or features that take data manipulation to the next level in terms of speed and ease.       
    
- `data.table::setkey()` allows users to order the data.table on specified columns which makes any form of computation on that involving keys insanely fast. It can lead to *2-3x* gains over regular data.table and *100-200x* gains over dplyr operations.     
      
- `melt()` and `cast()` to reshape data to wide or long format 
    
- joining (merging) large data.table(s) incredibly fast using a combination of `on` option and `setkeys()` function.  
      
- `fread()` to import large csv files lightning fast. It can also import files from a zipped file without unzipping it. (Insane!) 

---

class: inverse, center, middle
name: dplyr.data.table

# dplyr + data.table

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>


---
# dtplyr package

It is one of the most amazing packages out there which provides a data.table translation for dplyr code. It can be installed using `pacman::p_load(dtplyr)`.     
```{r, echo=FALSE}
pacman::p_load(dtplyr, tidyfast)
```

In order to use dtplyr, the dataframe object has to be converted to lazy data.table class using,
```{r, eval=FALSE}
gapminder_dtplyr = lazy_dt(gapminder::gapminder)
gapminder_dtplyr %>%
  filter(continent %in% c("Asia","Europe")) %>%
  group_by(continent, year) %>%
  summarize(mean_pop = mean(pop, na.rm=T), mean_gdppc = mean(gdpPercap, na.rm=T))
```

dtplyr automatically translates dplyr syntax to data.table equivalent in the backend. It is an effective way to take advantage of data.table without putting any effort into learning it's syntax.

---
# tidyfast package

It provides tidying functions built on data.table that also accepts `%>%` operator. The package can be installed using using `pacman::p_load(tidyfast)`. 


Transformation from long to wide.
```{r, eval=F}
dt = as.data.table(gapminder::gapminder) 
dt_wide = dt %>%
  dt_pivot_wider(names_from = year, values_from = c(lifeExp, pop, gdpPercap))
```

Transformation from wide to long
```{r, eval=F}
dt_wide %>%
  dt_pivot_longer(cols = c(-continent, -country), names_to = "type_year",
                  values_to = "values") %>%
  .[, c("type","year") :=  tstrsplit(type_year, "_", fixed=TRUE)] %>%
  .[, type_year := NULL] %>%
  dt_pivot_wider(names_from = type, values_from = values)
```


---
# some important points

- R is versatile and provides different functions to perform the similar task. It is the user's job to decide which is the appropriate function to use.            
    
- Do not be hell bent to fit every data wrangling exercise into a `dplyr` vs `data.table` problem. The correct way is to use both of them along with other packages depending on the requirement.        
    
- Remember whatever issue you're trying to solve has more often than not already been solved. Your job is to find the right answer. In other words, google and stackoverflow is your answer to every problem.     
    
- The easiest way to learn is to do. I believe it's better to straightaway jump into a task rather than read guides to no avail. Pick a task you already performed in Stata and try replicating in R. 

---
class: inverse, center, middle
name: thank.you

# thank you





```{r gen_pdf, include = FALSE, cache = FALSE, eval = FALSE, warning=FALSE}
pagedown::chrome_print("02_datawrangling.html", output = "02_datawrangling.pdf")
```






